Understanding the Nanoscale Structure of Inverted Hexagonal Phase Lyotropic Liquid Crystal Polymer Membranes

### 2018-10-11 Ben_Manuscripts/structure_paper/Final_Draft.tex
New toc graphic

  More highly selective nanoporous membranes would be extremely useful for
% BJC: in --> for
  performing complex aqueous separations with seawater and various types of
  wastewater. For example, Sodium chloride and boron in seawater
  \cite{fritzmann_state---art_2007} and organic micropollutants found in
  municipal and industrial wastewaters \cite{schwarzenbach_challenge_2006}
  represent just a few of the diverse contaminants of water sources. By
  efficiently separating contaminants from feed solutions with highly selective
  membranes, it is possible to reduce the number of required membrane passes and
  post-treatment steps needed for a given filtration process
  \cite{werber_materials_2016}, thus lowering energy requirements. Additionally,
  one can also extract valuable resources from the feed streams. For example,
  flowback water produced during hydraulic fracturing of shale formations
  contains dissolved species such as acetate whose extraction has economic value
  \cite{dischinger_application_2017}.

@@

  Ordered, nanostructured porous membrane materials have become an increasingly popular
  research area for aqueous separation applications because they offer the
  %BJC: separations --> separation 
  ability to control pore architecture at the molecular scale, thereby
  permitting the design of solute-specific separation membranes
  \cite{humplik_nanostructured_2011}. However, their development has been limited
  by the ability to synthesize and scale various technologies. Graphene sheets
  are atomically thick, which results in excellent water permeability but defects
  during manufacturing severely impact selectivity
  \cite{cohen-tanugi_multilayer_2016}.  Molecular dynamics (MD) simulations of
  carbon nanotubes have shown promise \cite{humplik_nanostructured_2011}, but
  synthetic techniques are unable to achieve scalable alignment and pore
  monodispersity \cite{hata_water-assisted_2004,maruyama_growth_2005}.
  Consequently, there is a need for new materials that could provide scalable
  nanostructured porous membranes.

@@

  \begin{figure}
	\centering
	\begin{subfigure}{.3\textwidth}
		\centering
		\includegraphics[width=\textwidth]{NaGA3C11.pdf}
		\caption{}~\label{fig:monomer}
	\end{subfigure}
	\begin{subfigure}{.3\textwidth}
		\centering
		\includegraphics[width=\textwidth]{monomer_twocolor.pdf}
		\caption{}~\label{fig:atomistic_monomer}
	\end{subfigure}
	\begin{subfigure}{0.3\linewidth}
		\centering
		\includegraphics[width=0.6\textwidth]{wedge_thick.pdf}
		\caption{}~\label{fig:wedge}
	\end{subfigure}
		\begin{subfigure}{0.4\linewidth}
		\centering
		\includegraphics[width=0.6\textwidth]{columns.pdf}
		\caption{}~\label{fig:wedge_layer}
	\end{subfigure}
	\begin{subfigure}{0.4\linewidth}
		\centering
		\includegraphics[width=0.6\textwidth]{hexagonal_packing.pdf}
		\caption{}~\label{fig:hex_packing_simple}
	\end{subfigure}
	\caption{(a) The LLC monomer Na-GA3C11 (b) rendered atomistically (c)
	exhibits wedge-like character. (d) Monomers stack on top of each other to
	create columns with short range order, then assemble into pores with 
        %BJC: order --> order, (added comma) 
	hydrophilic head groups (blue) facing towards the pore center. (e) The 
	pores assemble into hexagonally packed columnar mesophases.}~\label{fig:assembly}
  \end{figure}

@@

  In this study, we achieve a more realistic atomistic description of a
  %BJC: of --> of a
  hexagonal LLC polymer membranes than, to our knowledge, has ever previously
  been created, and explore what new structural information can be gained and
  what structure hypotheses are supported by this model. We validate the results
  using as much experimental information as possible. We are most interested in
  reproducing the conclusions about structure drawn from small-angle X-ray
  scattering (SAXS) and wide-angle X-ray scattering (WAXS) experiments, as well
  as in matching ionic conductivity measurements \cite{feng_thin_2016}.

@@

	\item \textit{R-spots}: Within R-alkanes, are four spots of higher
	relative intensity.  Accordingly, we name these reflections R-spots. The
	location of all spots is $\sim37^{\circ}$ from the $q_r$ axis in their
	%BJC: q_z --> q_r
	respective quadrants. In many liquid crystal systems, such spots are explained
	as the result of alkane chains tilted with respect to the membrane
	plane\cite{govind_simple_2001}.

@@

  The time scale for self-assembly of monomers into the hexagonal phase is
  unknown and likely outside of a reasonable length for an atomistic simulation,
  requiring a more efficient way to build the system. Previous work has shown
  that a united-atom model of an LLC can self-assemble into the
  %BJC: show a united-atom --> that a united-atom ; of LLC --> of an LLC
  H\textsubscript{II} phase configuration in $\sim$1000 ns
  \cite{mondal_self-assembly_2013}. We attempted atomistic self-assembly by
  packing monomers into a box using Packmol \cite{martinez_packmol:_2009}.
  Simulations of greater than 100 ns show no indicators of progress towards an
  ordered system (see Section \ref{S-section:self_assembly} of the SI). To bypass
  the slow self-assembly process, we used automated procedures to
  %BJC: a variety of automated --> automated
  assemble monomers into a structure close to one of a number of hypothesized
  equilibrium configurations (See Figure \ref{S-fig:build_procedure} of the SI).
  These initial structures may result in equilibration into slowly-intercoverting
  metastable states, an issue we will address later in the paper.

@@

  We gave the simplified models the same amount of disorder present in the
  atomistic simulations. We observe two sources of disorder in our atomistic
  simulations: thermal motion of atoms during the simulation and quenched
  disorder created by rapid structural rearrangement during early equilibration
  that is largely locked into place for the remainder of the simulation, even at
  %BJC: and --> that
  the 100 ns time scale. We measured thermal disorder by calculating the standard
  deviation of the distribution of head group COM positions from their average
  positions. We measured quenched disorder by calculating the standard deviation
  of the distribution of head group COMs from their idealized average positions.
  In the $z$-direction, we measured the deviation of the head group COMs from an
  equally spaced column of head groups. In the $xy$ plane, we calculated the
  standard deviation in radial position from the pore center and the angular
  deviation from equally spaced points surrounding the pore center. This method
  for calculating quenched disorder inherently includes the convoluted thermal
  disorder.

@@

  Systems built with 5 columns around each pore equilibrate to a pore spacing
  that is most consistent with the experimental value of 4.12 nm derived from
  SAXS measurements (Figure~\ref{fig:SAXS}). Ordered basin systems built with 4
  columns per pore equilibrate to an average pore spacing 0.25 nm lower than
  experiment. Ordered basin systems built with 6 columns per pore, have an
  equilibrated pore spacing ca. 0.50 nm higher than experiment. Monomers in
  disordered basin systems built with 6 columns-per-pore agree with experimental
  pore-to-pore distances within error, but stack too far apart. 6
  %BJC: Monomers in 6 --> 6
  column per pore disordered sandwiched and disordered parallel displaced
  configurations stack $\sim$~4.87 and 4.94 \AA~apart respectively, which is
  $\sim$ 1.2~\AA~further apart than suggested by experiment. 5 column-per-pore
  systems stack, at a maximum, 0.9 \AA~further apart than experiment (see
  Table~\ref{table:correlation_length}). The remainder of this discussion will
  therefore focus on the analysis of systems built with 5 columns per pore.

@@

  However, when we allowed monomer columns to move independently in the
  $z$-direction, the intensity of R-$\pi$ drops and the $q_r$ cross-section of
  the simulated diffraction patterns smooths out. We can demostrate this by
  creating a simplified sandwiched configuration (as described above), with
  varying amounts of statistical correlation between columns (see
  Figure~\ref{fig:column_displacement}). In the low independent motion limit, the
  COM of columns are situated at the same $z$-coordinate.  We allow increasing
  independence of columns by randomly shifting each column in the $z$-direction
  according to a uniform distribution bounded by (0, $f \times \mathit{d}$) where
  $f$ ranges between 0 and 1 and $d$ is the vertical distance between scatterers.
  %BJC: removed this sentence since I do parallel displaced config in the ensemble
  %We limit this discussion to the sandwiched configuration since it is not
  %feasible for columns in the parallel displaced configuration, as we have set
  %them up, to move past each other. 
  When $f = 1$, the columns move independently and the $q_r$ cross-section
  smooths out completely (Figure~\ref{fig:sf_qy_sr100}). Any amount of dependence
  results in some presence of sharp peaks, though significantly attenuated at $f$
  nearer 1. The intensity of R-$\pi$ also decreases with increasing column
  independence. We see an 9-fold decrease in the intensity of R-$\pi$ when $f=1$
  versus when $f=0$ for our simplified model systems.

@@

  \begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{z_correlation_fullbox.pdf}
  \caption{When we average all $z$-slices of the 3D correlation function, the 
  correlation length of the ordered sandwich configuration nearly triples compared 
  %BJC: added figure reference below. Check that it works
  to that shown in Figure~\ref{fig:z_correlation_sandwich}.}\label{fig:z_correlation_fullbox}
  \end{figure}

@@

  The appearance of R-double implies a vertical modulation in electron density
  every 7.4 \AA. We are not able to achieve such modulation using our simple
  initial configurations. Although the position of monomers in parallel displaced
  configurations alternate every other layer, such a configuration will not
  produce R-double, but only off-axis reflections at the same $q_z$
  value~\cite{harburn_atlas_1975}. There is not a unique solution that describes
  the origin of R-double. Extracting the exact relationship between a diffraction
  pattern and its real space configuration is well-known as the phase
  problem~\cite{taylor_phase_2003}. We have proposed configurations that
  result in the appearance of R-double below, and we can speculate which makes
  the most physical sense. 
  %BJC: seems redundant from last sentence of above paragraph
%It appears that we cannot achieve a long-term stable
%  system that exhibits R-double without the inclusion of small amounts of water
%  in the pores.

>>>>

  More highly selective nanoporous membranes would be extremely useful for
  performing complex aqueous separations with seawater and various types of
  wastewater. For example, Sodium chloride and boron in seawater
  \cite{fritzmann_state---art_2007} and organic micropollutants found in
  municipal and industrial wastewaters \cite{schwarzenbach_challenge_2006}
  represent just a few of the diverse contaminants of water sources. By
  efficiently separating contaminants from feed solutions with highly selective
  membranes, it is possible to reduce the number of required membrane passes and
  post-treatment steps needed for a given filtration process
  \cite{werber_materials_2016}, thus lowering energy requirements. Additionally,
  one can also extract valuable resources from the feed streams. For example,
  flowback water produced during hydraulic fracturing of shale formations
  contains dissolved species such as acetate whose extraction has economic value
  \cite{dischinger_application_2017}.

@@

  Ordered, nanostructured porous membrane materials have become an increasingly popular
  research area for aqueous separation applications because they offer the
  ability to control pore architecture at the molecular scale, thereby
  permitting the design of solute-specific separation membranes
  \cite{humplik_nanostructured_2011}. However, their development has been limited
  by the ability to synthesize and scale various technologies. Graphene sheets
  are atomically thick, which results in excellent water permeability but defects
  during manufacturing severely impact selectivity
  \cite{cohen-tanugi_multilayer_2016}.  Molecular dynamics (MD) simulations of
  carbon nanotubes have shown promise \cite{humplik_nanostructured_2011}, but
  synthetic techniques are unable to achieve scalable alignment and pore
  monodispersity \cite{hata_water-assisted_2004,maruyama_growth_2005}.
  Consequently, there is a need for new materials that could provide scalable
  nanostructured porous membranes.

@@

  \begin{figure}
	\centering
	\begin{subfigure}{.3\textwidth}
		\centering
		\includegraphics[width=\textwidth]{NaGA3C11.pdf}
		\caption{}~\label{fig:monomer}
	\end{subfigure}
	\begin{subfigure}{.3\textwidth}
		\centering
		\includegraphics[width=\textwidth]{monomer_twocolor.pdf}
		\caption{}~\label{fig:atomistic_monomer}
	\end{subfigure}
	\begin{subfigure}{0.3\linewidth}
		\centering
		\includegraphics[width=0.6\textwidth]{wedge_thick.pdf}
		\caption{}~\label{fig:wedge}
	\end{subfigure}
		\begin{subfigure}{0.4\linewidth}
		\centering
		\includegraphics[width=0.6\textwidth]{columns.pdf}
		\caption{}~\label{fig:wedge_layer}
	\end{subfigure}
	\begin{subfigure}{0.4\linewidth}
		\centering
		\includegraphics[width=0.6\textwidth]{hexagonal_packing.pdf}
		\caption{}~\label{fig:hex_packing_simple}
	\end{subfigure}
	\caption{(a) The LLC monomer Na-GA3C11 (b) rendered atomistically (c)
	exhibits wedge-like character. (d) Monomers stack on top of each other to
	create columns with short range order, then assemble into pores with 
	hydrophilic head groups (blue) facing towards the pore center. (e) The 
	pores assemble into hexagonally packed columnar mesophases.}~\label{fig:assembly}
  \end{figure}

@@

  In this study, we achieve a more realistic atomistic description of
  hexagonal LLC polymer membranes than, to our knowledge, has ever previously
  been created, and explore what new structural information can be gained and
  what structure hypotheses are supported by this model. We validate the results
  using as much experimental information as possible. We are most interested in
  reproducing the conclusions about structure drawn from small-angle X-ray
  scattering (SAXS) and wide-angle X-ray scattering (WAXS) experiments, as well
  as in matching ionic conductivity measurements \cite{feng_thin_2016}.

@@

	\item \textit{R-spots}: Within R-alkanes, are four spots of higher
	relative intensity.  Accordingly, we name these reflections R-spots. The
	location of all spots is $\sim37^{\circ}$ from the $q_r$ axis in their
	respective quadrants. In many liquid crystal systems, such spots are explained
	as the result of alkane chains tilted with respect to the membrane
	plane\cite{govind_simple_2001}.

@@

  The time scale for self-assembly of monomers into the hexagonal phase is
  unknown and likely outside of a reasonable length for an atomistic simulation,
  requiring a more efficient way to build the system. Previous work has shown
  that a united-atom model of an LLC can self-assemble into the
  H\textsubscript{II} phase configuration in $\sim$1000 ns
  \cite{mondal_self-assembly_2013}. We attempted atomistic self-assembly by
  packing monomers into a box using Packmol \cite{martinez_packmol:_2009}.
  Simulations of greater than 100 ns show no indicators of progress towards an
  ordered system (see Section \ref{S-section:self_assembly} of the SI). To bypass
  the slow self-assembly process, we used automated procedures to
  assemble monomers into a structure close to one of a number of hypothesized
  equilibrium configurations (See Figure \ref{S-fig:build_procedure} of the SI).
  These initial structures may result in equilibration into slowly-intercoverting
  metastable states, an issue we will address later in the paper.

@@

  We gave the simplified models the same amount of disorder present in the
  atomistic simulations. We observe two sources of disorder in our atomistic
  simulations: thermal motion of atoms during the simulation and quenched
  disorder created by rapid structural rearrangement during early equilibration
  that is largely locked into place for the remainder of the simulation, even at
  the 100 ns time scale. We measured thermal disorder by calculating the standard
  deviation of the distribution of head group COM positions from their average
  positions. We measured quenched disorder by calculating the standard deviation
  of the distribution of head group COMs from their idealized average positions.
  In the $z$-direction, we measured the deviation of the head group COMs from an
  equally spaced column of head groups. In the $xy$ plane, we calculated the
  standard deviation in radial position from the pore center and the angular
  deviation from equally spaced points surrounding the pore center. This method
  for calculating quenched disorder inherently includes the convoluted thermal
  disorder.

@@

  Systems built with 5 columns around each pore equilibrate to a pore spacing
  that is most consistent with the experimental value of 4.12 nm derived from
  SAXS measurements (Figure~\ref{fig:SAXS}). Ordered basin systems built with 4
  columns per pore equilibrate to an average pore spacing 0.25 nm lower than
  experiment. Ordered basin systems built with 6 columns per pore, have an
  equilibrated pore spacing ca. 0.50 nm higher than experiment. Monomers in
  disordered basin systems built with 6 columns-per-pore agree with experimental
  pore-to-pore distances within error, but stack too far apart. 6
  column per pore disordered sandwiched and disordered parallel displaced
  configurations stack $\sim$~4.87 and 4.94 \AA~apart respectively, which is
  $\sim$ 1.2~\AA~further apart than suggested by experiment. 5 column-per-pore
  systems stack, at a maximum, 0.9 \AA~further apart than experiment (see
  Table~\ref{table:correlation_length}). The remainder of this discussion will
  therefore focus on the analysis of systems built with 5 columns per pore.

@@

  However, when we allowed monomer columns to move independently in the
  $z$-direction, the intensity of R-$\pi$ drops and the $q_r$ cross-section of
  the simulated diffraction patterns smooths out. We can demostrate this by
  creating a simplified sandwiched configuration (as described above), with
  varying amounts of statistical correlation between columns (see
  Figure~\ref{fig:column_displacement}). In the low independent motion limit, the
  COM of columns are situated at the same $z$-coordinate.  We allow increasing
  independence of columns by randomly shifting each column in the $z$-direction
  according to a uniform distribution bounded by (0, $f \times \mathit{d}$) where
  $f$ ranges between 0 and 1 and $d$ is the vertical distance between scatterers.
  When $f = 1$, the columns move independently and the $q_r$ cross-section
  smooths out completely (Figure~\ref{fig:sf_qy_sr100}). Any amount of dependence
  results in some presence of sharp peaks, though significantly attenuated at $f$
  nearer 1. The intensity of R-$\pi$ also decreases with increasing column
  independence. We see an 9-fold decrease in the intensity of R-$\pi$ when $f=1$
  versus when $f=0$ for our simplified model systems.

@@

  \begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{z_correlation_fullbox.pdf}
  \caption{When we average all $z$-slices of the 3D correlation function, the 
  correlation length of the ordered sandwich configuration nearly triples compared 
  to that shown in Figure~\ref{fig:z_correlation_sandwich}.}\label{fig:z_correlation_fullbox}
  \end{figure}

@@

  The appearance of R-double implies a vertical modulation in electron density
  every 7.4 \AA. We are not able to achieve such modulation using our simple
  initial configurations. Although the position of monomers in parallel displaced
  configurations alternate every other layer, such a configuration will not
  produce R-double, but only off-axis reflections at the same $q_z$
  value~\cite{harburn_atlas_1975}. There is not a unique solution that describes
  the origin of R-double. Extracting the exact relationship between a diffraction
  pattern and its real space configuration is well-known as the phase
  problem~\cite{taylor_phase_2003}. We have proposed configurations that
  result in the appearance of R-double below, and we can speculate which makes
  the most physical sense.
